---
title: "assignment-8"
author: "kimp"
date: "12/12/2017"
output: html_document
---

We will apply the K-means clustering algorithm to the acceleration data we collected about Wii Bowling Arm Swings using the Sensor Kinetics App.

"The k-means clustering algorithm classifies n points into k clusters by assigning each point to the cluster whose average value on a set of p variables is nearest to it by some distance measure (usually Euclidean) on that set. The algorithm computes these assignments iteratively, until reassigning points and recomputing averages (over all points in a cluster) produces no changes" (https://www.cs.uic.edu/~wilkinson/Applets/cluster.html).

First, I  install the three packages: "dplyr", "tidyr", and "ggplot2". 

```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
```

Next, I combine the aggregated data from all 5 participants into one master dataframe. 

```{r}

setwd("~/Desktop/HUDK4050/assignment-8/Assignment_8_Data/Clean_Data")
file.names <- list.files()
DY_data <- read.csv("dy-clean data with scores.csv")
KP_data <- read.csv("KPmaster.csv")
NA_data <- read.csv("NA_master.csv")
TZ_data <- read.csv("tz_acc_combined.csv")
ZY_data <- read.csv("ZY_master.csv")
all_data <- rbind(DY_data, KP_data, NA_data, TZ_data, ZY_data)
str(all_data)
summary(all_data)
  
```

"The algorithm will treat each row as a value belonging to a person" so we need to make each row descriptive of one trial. I will do this by grouping by name, filtering by each trial attempt (1-10), and summarize the mean of the XYZ coordinates of the acceleration.

As a result, I should total 50 rows to equal the 10 attempts of 5 participants.

Note: I have also taken the absolute value of each accleration prior to calculating the mean.

```{r}
x_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_1 <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data1 <- bind_cols(x_1, y_1, z_1)
sum_data1$Trial <- 1
Score <- all_data %>% group_by(Name) %>% filter(Trial == 1) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data1 <- bind_cols(sum_data1, Score)
```

```{r}
x_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_2 <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data2 <- bind_cols(x_2, y_2, z_2)
sum_data2$Trial <- 2
score <- all_data %>% group_by(Name) %>% filter(Trial == 2) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data2 <- bind_cols(sum_data2, Score)
```

```{r}
x_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_3 <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data3 <- bind_cols(x_3, y_3, z_3)
sum_data3$Trial <- 3
score <- all_data %>% group_by(Name) %>% filter(Trial == 3) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data3 <- bind_cols(sum_data3, Score)
```

```{r}
x_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_X_acc = mean(abs(acc_X_value))) 
y_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_4 <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data4 <- bind_cols(x_4, y_4, z_4)
sum_data4$Trial <- 4
score <- all_data %>% group_by(Name) %>% filter(Trial == 4) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data4 <- bind_cols(sum_data4, Score)
```

```{r}
x_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_5 <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data5 <- bind_cols(x_5, y_5, z_5)
sum_data5$Trial <- 5
score <- all_data %>% group_by(Name) %>% filter(Trial == 5) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data5 <- bind_cols(sum_data5, Score)
```

```{r}
x_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_6 <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data6 <- bind_cols(x_6, y_6, z_6)
sum_data6$Trial <- 6
score <- all_data %>% group_by(Name) %>% filter(Trial == 6) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data6 <- bind_cols(sum_data6, Score)
```

```{r}
x_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_7 <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data7 <- bind_cols(x_7, y_7, z_7)
sum_data7$Trial <- 7
score <- all_data %>% group_by(Name) %>% filter(Trial == 7) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data7 <- bind_cols(sum_data7, Score)
```

```{r}
x_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_8 <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data8 <- bind_cols(x_8, y_8, z_8)
sum_data8$Trial <- 8
score <- all_data %>% group_by(Name) %>% filter(Trial == 8) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data8 <- bind_cols(sum_data8, Score)
```

```{r}
x_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_9 <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data9 <- bind_cols(x_9, y_9, z_9)
sum_data9$Trial <- 9
score <- all_data %>% group_by(Name) %>% filter(Trial == 9) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data9 <- bind_cols(sum_data9, Score)
```

```{r}
x_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_X_acc = mean(abs(acc_X_value)))
y_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_Y_acc = mean(abs(acc_Y_value))) %>% select(-Name)
z_10 <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(avg_Z_acc = mean(abs(acc_Z_value))) %>% select(-Name)
sum_data10 <- bind_cols(x_10, y_10, z_10) 
sum_data10$Trial <- 10
score <- all_data %>% group_by(Name) %>% filter(Trial == 10) %>% summarise(Score = mean(Score)) %>% select(-Name)
sum_data10 <- bind_cols(sum_data10, Score)
```

Finally, I will combine all dataframes into one dataframe, sum_data.

```{r}
sum_data <- bind_rows(sum_data1, sum_data2, sum_data3, sum_data4, sum_data5, sum_data6, sum_data7, sum_data8, sum_data9, sum_data10)
```

## Part I

I will create clusters by the XYZ acclerations and interpret the clusters by observing the clusters along a X-Y plot of score and acceleration, XYZ.

First, I scale only the acceleration XYZ data.

Logic:
"Another pre-processing step used in K-means is to standardize the values so that they have the same range. We do this because we want to treat each week as equally important - if we do not standardise then the week with the largest range will have the greatest impact on which clusters are formed. We standardise the values by using the "scales()" command."
* Copied and pasted from HUDK4050 Assignment 3.

```{r}
scale_acc <- select(sum_data, avg_X_acc, avg_Y_acc, avg_Z_acc)
scale_acc <- scale(scale_acc)
scale_acc <- as.data.frame(scale_acc)
```

```{r}
sum_data <- select(sum_data, -avg_X_acc, -avg_Y_acc, -avg_Z_acc)
sum_data <- bind_cols(sum_data,scale_acc)
```


"Now we will run the K-means clustering algorithm we talked about in class. 
1) The algorithm starts by randomly choosing some starting values 
2) Associates all observations near to those values with them
3) Calculates the mean of those clusters of values
4) Selects the observation closest to the mean of the cluster
5) Re-associates all observations closest to this observation
6) Continues this process until the clusters are no longer changing"

We will use 3 clusters. 

```{r}
fit <- kmeans(scale_acc, 3)
sum_data$cluster <- fit$cluster
```

## Part II - Visualization

```{r}
sum_data$cluster <- as.factor(sum_data$cluster)
```

Visualization for avg XYZ accleration per trial by each participant. 

```{r}
ggplot(sum_data, aes(Trial, avg_X_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average X Acceleration") + ggtitle("Avg X Acceleration per Trial by Each Participant")

ggplot(sum_data, aes(Trial, avg_Y_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average Y Acceleration") + ggtitle("Avg Y Acceleration per Trial by Each Participant")

ggplot(sum_data, aes(Trial, avg_Z_acc, colour = Name)) + geom_line(stat="identity") + xlab("Trial") + ylab("Average Z Acceleration") + ggtitle("Avg Z Acceleration per Trial by Each Participant")
```

Visualization for avg XYZ acceleration compared to scoreby clusters.

Note: I have removed observed outliers for X and Y accelerations to improve visualization of the clusters. 

```{r}
x <- filter(sum_data, avg_X_acc < 3)
y <- filter(sum_data, avg_Y_acc > -1.75)

ggplot(x, aes(avg_X_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average X Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average X Acceleration") + theme(plot.title = element_text(hjust = 0.5))

ggplot(y, aes(avg_Y_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average Y Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average Y Acceleration") + theme(plot.title = element_text(hjust = 0.5))

ggplot(sum_data, aes(avg_Z_acc, Score, colour = cluster)) + geom_jitter(stat="identity") + xlab("Average Z Acceleration") + ylab("Score") + geom_smooth(method = 'lm', se = FALSE) + ggtitle("Clusters by Score and Average Z Acceleration") + theme(plot.title = element_text(hjust = 0.5))
```

## Part 3 - Cluster Interpretation

From the visualizations, I'm seeing the following:

Cluster 1 - positive x, y, z acceleration, AND scores > 6.
Cluster 2 - negative x, y, positive z, AND scores > 6 and scores < 8.
Cluster 3 - positive and negative x (range from -2,2), y ~= 0, negative z, AND scores > 9 and scores < 4 (with 1 outlier score around 7).